An Million Transcriptome Ocular Meta-Atlas for Discovery and Development

### 2020-11-05 manuscript_00.Rmd
text updates

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "The development of highly scalable single cell transcriptome technology has allowed for a high resolution view into huge numbers of single cell transcriptomes. Analyzing the transcriptomes between different projects is highly desirable as this would theoretically allow for better assessment of which effects are consistent across independent studies. However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation. While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best. We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing methods and parameters. We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 studies and three species to create a massive ocular single cell transcriptome meta-atlas. This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
---

@@

methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none', 'liger'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','liger','magic','scanorama','scArches','scVI', 'none')) 
```

# Introduction
## Many published ocular atlases

The recent explosion of research using single cell transcriptome technology has led to insights in development, disease progression, etc. Some of the seminal work in the single cell transcriptomics field (Macosko et al) have used the retina. For example in the ocular community, over a dozen studies have used single cell technology to profile cell type patterns, etc.

DISCUSS RETINA DEVELOPMENT/CELL TYPE stuff

## Data not easily available for all

While these groups has leveraged this new technology to [do stuff], there is a large chasm to cross the enable outside groups to use the data in new ways. Several of these papers are accompanied by reactive web apps, which allow for relatively quick checks of gene expression across cell type or cluster assignment. However these web apps have minimal to no data exploration tools and are slow to operate, especially with the larger datasets.

## Impossible to compare results across independent studies
Most crucially, these paper data web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study. Thus it is impossible to compare directly across datasets. When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.

While lightly processed counts data is often provided, differences in both single cell technology (e.g. droplet or well based) and bioinformatic processing choices make it impractical to quickly aggregate the datasets together as differences found may be technical.

## There is a path 
As the raw sequence data is required to be deposited on publication, it is possible to remove the technical differences due to bioinformatic processing choices by re-processing all of the data in a unified framework, though this still requires a large amount of bioinformatic expertise and compute infrastructure. Also there still remain technical (or batch) effects due to the variety in single cell technologies available and variation in tissue handling and processing across different scientific groups. Fortunately the single cell community has recognized that removal of technical (also referred to as batch) effects is a crucial issue and have independently developed many tools.

@@

<!-- ## As Single cell datasets getting larger, crucial for algorithms to be highly performant -->
<!--   - Valentine Svensson figure? -->
<!--   - MOCA atlas (mouse 2e6 cells?) -->

<!-- Our multidimensional atlas of 1e6 cells will be a useful reference dataset in: -->
<!--   - batch effect correction tools -->
<!--   - clustering algorithms -->
<!--   - cell type prediction / projection -->
<!--   - trajectory inference -->

@@

We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology. We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r supTab_cap(name = 'study_cell_count', display = 'cite')`). 
  To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project. In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files and re-extracted the fastq. After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.

```{r fig1, fig.width=13, fig.height=5, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
fig1_cap <- fig_cap(name='fig1', caption =  'a. Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism. b. Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods to produce batch corrected latent dimensions (LDims), then downstream analysis outputs like clustering, differential gene testing (DiffT), and 2D UMAP visualization. c. Cell type counts extracted from published studies for select cell types, split by species. Count of study accessions for each species overlaid on bar plot.')

@@

## `r cell_info_labels %>% nrow() %>% as.integer() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control

  Gene-level counts were quantified with the kallisto bustool pseudo-aligner. For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R. After empty droplet removal, we had `r format((umap %>% nrow() %>% as.numeric()) - (meta_filter %>% nrow() %>% as.numeric()), big.mark=",", scientific=FALSE)` cells. We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts. For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets. After these standard quality control steps we were left with `r format(umap %>% nrow, big.mark = ",", scientific=FALSE)` cells.

@@

  A core objective of many scRNA based studies is labeling distinct cell types. As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence. After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types.

## Integration benchmarking of `r methods %>% length()` tools

  A key step in extracting biological information out of the unified dataset is to remove the technical effects from the data. A wide variety of tools have been designed for this function. As we were uncertain which would perform the best and it is difficult to manually assess performance we designed a R-based benchmarking system to evaluate performance (scPOP). There are two key metrics which have to be balanced in order to optimize performance: batch mixing (e.g. different studies with the same cell type should be similar) and cell type or cluster purity (where different cell types or clusters should be distinct). While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify this process.
  We define batch as being each biological sample. We assume each study is at least one unique sample. We also studied the metadata and methods of each study to determine if the particular study contained multiple samples. In the end we identified `r meta_filter$batch %>% table() %>% length()` batches to be integrated. 
  The Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, were used to assess integration performance. We also use the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB. The LISI and ASW were used to measure batch mixing (where lower is better), celltype mixing (higher is better), and cluster mixing (higher and better). NMI and ARI were used to assess the consistency of celltype to cluster assignment (where 1 is perfect correspondence between cluster and cell type). The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space. 
  To visualize the interplay between batch mixing and cell type distinction we plot in FIG XXXXX the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cell type LISI on the x-axis (higher is better). The best performer on both metrics will be in the top right corner.

@@

  To rank the integration algorithms across the various benchmarks, each score was Z scale normalized so that each score has an equal weight. The scores were then summed which allows us to rank the performance of each integration tool across the benchmarks. scVI has the strongest performance, with a sum score of SOMETHING. The next highest performers are Harmony, fastMNN, and Scanorama.

@@

## xgboost model built to label unknown cell types
  To further study cell type specific expression patterning we needed to label the `r format(umap %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells. Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters. However, as we had hundreds of thousands of pre-labelled cells we decided to leverage the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details). The model was used to predict the cell type assignments for all cells in scEiaD. In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct the small number of mislabels in the truth set.

@@

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "The development of highly scalable single cell transcriptome technology has allowed for a high resolution view into huge numbers of single cell transcriptomes. Analyzing the transcriptomes between different projects is highly desirable as this would theoretically allow for better assessment of which effects are consistent across independent studies. However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation. While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best. We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations. We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 datsets and three species to create a massive ocular single cell transcriptome meta-atlas. This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
---

@@

methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none', 'liger'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','liger','magic','scanorama','scArches','scVI', 'none'))

qc <- readr::read_tsv('~/data/massive_integrated_eye_scRNA/QC.tsv.gz')
```

# Introduction
## Explosion of single cell transcriptomic atlases
The  recent introduction of relatively low cost and high throughput single cell sequencing technlogy has led to an explosion of research across many fields. GIVE SOME EXAMPLES. As of 2020, over 30 million cells have been sequenced across n studies (Svennson Pachter http://www.nxn.se/single-cell-studies)

## Many in the eye 
  Some of the seminal work in the high throughput single cell transcriptomics field (Macosko et al) have used the retina. As of late 2020, over a dozen published studies have used single cell technology to profile cell type specific gene expression patterns, developmental trends, species specific differentiation, and disease alterations.

## Impossible to easily compare results across independent studies
Study specific web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study. There are notable efforts to create web portals (UCSC Single Cell Browser, EBI Single Cell Expression Atlas) that allow for GUI-based viewing of data for most publicly available datasets. However, these again have the shortcoming as the study-specific web portals discussed above in that you still cannot directly compare between independent studies. In these situations it is impossible to compare directly across datasets. When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.

<!-- ## Aggregation of results to create meta-atlas important -->

<!-- While these groups has leveraged this new technology to [do stuff], there is a large chasm to cross the enable outside groups to use the data in new ways. Several of these papers are accompanied by reactive web apps, which allow for relatively quick checks of gene expression across cell type or cluster assignment. However these web apps have minimal to no data exploration tools and are slow to operate, especially with the larger datasets.  -->

<!-- While lightly processed counts data is often provided, differences in both single cell technology (e.g. droplet or well based) and bioinformatic processing choices make it impossible to simply concatenate the datasets together as differences found will likely be technical.  -->

## There is a path 
As the raw sequence data is required to be deposited on publication, it is possible to remove the technical differences due to bioinformatic processing choices by re-processing all of the data in a unified framework, though this still requires a large amount of bioinformatic expertise and compute infrastructure. Also there still remain technical (or batch) effects due to the variety in single cell technologies available and variation in tissue handling and processing across different scientific groups. Fortunately the single cell community has recognized that removal of technical (also referred to as batch) effects is a crucial issue and have independently developed many tools. The challenge is still to determine which tool works best.

## scRNA datasets are rapidly increasing in size
  - Valentine Svensson figure?
  - http://www.nxn.se/single-cell-studies
  - MOCA atlas (mouse 2e6 cells)

@@

@@

  The first step in building a meta-atlas is identifying studies to draw the data from. We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and then hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology. We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`). 
  To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project. In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files and re-extracted the fastq. After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.

```{r fig1, fig.width=13, fig.height=5, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
fig1_cap <- fig_cap(name='fig1', caption =  'a. Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism. b. Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization. c. Cell type counts extracted from published studies for the more common retina cell types, split by species. Count of study accessions for each species overlaid on bar plot.')

@@

## `r cell_info_labels %>% nrow() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control

  Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples. For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R. After empty droplet removal, we had `r format(cell_info_labels %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells. We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts. For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets. After these standard quality control steps we were left with `r format(umap %>% nrow, big.mark = ",", scientific=FALSE)` cells.

@@

  A core objective of many scRNA based studies is labeling distinct cell types. As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence. After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1', display = 'cite')`).

## Running `r methods %>% length()` tools in a Snakefile-based system

  A key step in extracting biological information out of the unified dataset is to remove the technical effects from the data. A wide variety of tools have been designed for this function. As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that effect integration: 2000 hyper variable genes, the louvain clustering algorithm with 7 k-nearest neighbors

  There are two key metrics which have to be balanced in order to optimize performance: batch mixing (e.g. different studies with the same cell type should be similar) and cell type or cluster purity (where different cell types or clusters should be distinct). While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify this process.
  We define batch as being each biological sample. We assume each study is at least one unique sample. We also studied the metadata and methods of each study to determine if the particular study contained multiple samples. In the end we identified `r meta_filter$batch %>% table() %>% length()` batches to be integrated.

# scPOP wraps several different methods at measuring integration performance.

  Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, are used as well as the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB. The LISI and ASW were used to measure batch mixing (where lower is better), celltype mixing (higher is better), and cluster mixing (higher and better). NMI and ARI were used to assess the consistency of celltype to cluster assignment (where 1 is perfect correspondence between cluster and cell type). The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space. 
  To visualize the interplay between batch mixing and cell type distinction we plot in FIG XXXXX the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cell type LISI on the x-axis (higher is better). The best performer on both metrics will be in the top right corner.

@@

  To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight. scPOP then sums them all which allows us to rank the performance of each integration tool across the benchmarks. For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or sepraration. scPOP finds that scVI has the strongest performance, with a sum score of SOMETHING. The next highest performers are Harmony, fastMNN, and Scanorama. We do not claim that these results are gegeral - rather we emphasize that for each batch correction analysis that scPOP or some similar approach be used to identify the optimal method for each dataset.

@@

## xgboost model built to label unknown cell types
  To further study cell type specific expression patterning we needed to label the `r format(umap %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells. Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters. However, as we had hundreds of thousands of pre-labelled cells we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input. The trained model was used to predict the cell type assignments for all cells in scEiaD. In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set.

@@

# Supplemental
```{r supFig1, fig.width=4, fig.height=7, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination. The right bar specifies the number of each cells in each post QC category. In silico doublets were identified with scrublet and DoubletDetector.')

source('figs_and_tables/sup_fig1__qc_failure.R')
supFig1_plot
```

```{r supTab1, fig.width=4, fig.height=7, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
supTab1_cap <- supTab_cap(name='supTab1_cap', caption =  'Counts for cell type labels. Published are the author created labels from the published datasets. Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')

source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
ctTable
```
